Haplotype Inference via Boolean Satisfiability and Integer Linear Programming

### 2020-12-10 paper.tex
final modifications

\documentclass[12pt,a4paper]{article}
%\usepackage[margin=1in]{geometry}

@@

\section{Introduction}
Haplotype inference is a crucial problem of modern genetics with wide-ranging applications in, e.g., correlating disease susceptibility to individuals' haplotypes.
Due to technical difficulties the direct measurement of haplotypes is challenging, and current methods rely on indirectly inferring haplotypes from genotype data.

@@

In practice the input formula is usually restricted to be in \emph{conjunctive normal form} (CNF).
A formula $\phi$ is in CNF if it is a conjunction of \emph{clauses}
$$ \bigwedge_{i\in I} C_i $$
where each clause $C_i$ is a disjunction of \emph{literals}
$$ C_i = \bigvee_{j\in J_i} l_j $$
and each literal is either an atom $p$ or a negated atom $\neg p$.
CNF is a convenient standard format partly due to the fact that efficient polynomial-time algorithms exist for transforming arbitrary formulas into \emph{linear-size, equisatisfiable} CNF formulas~\cite{tseitin1983, DBLP:journals/jsc/PlaistedG86}.

@@

\section{Haplotype inference}
\label{sec:hipp}
\TODO{How relevant is the 2-allele haplotype inference? How to generalize to several alleles?}

The human genetic information is contained in the double-stranded DNA molecules consisting of a linear sequence of 4 bases (\textbf{A}denosine, \textbf{C}ytosine, \textbf{G}uanine, \textbf{T}hymine) in the phosphate-deoxyribose backbone.
The DNA molecules, which are packed into \emph{chromosomes}, consist of protein-coding intervals, which are called genes, and other non-coding parts.
Different variants of a gene are called \emph{alleles}.
Humans are a \emph{diploid} species, which means that we have two copies of each chromosome (one from each parent).
We thus have two alleles of each gene, and we are said to be \emph{homozygous} with regard to a gene if we have two identical alleles and \emph{heterozygous} otherwise.

\emph{Haplotype} refers to the information contained in a single chromosome whereas the conflated information contained in the pair of  homologous chromosomes inherited is referred to as the \emph{genotype}.
In practice it is difficult and expensive to experimentally detect haplotype information.
Detecting genotype information has, however, become much faster, cheaper and more reliable in the recent decades.
Potential haplotypes can be computationally inferred from genotype data, and this is the problem referred to as the haplotype inference (HI) problem.
Haplotype data has the potential to yield new insight into genetic predisposition to, e.g., diseases, which makes the HI problem of great importance.

@@

\TODO{How does $n$ affect the accuracy?}

\TODO{Elaborate: "relatively accurate haplotype data"}

This condition has been experimentally validated to yield relatively accurate haplotype data~\cite{DBLP:journals/bioinformatics/WangX03a}.
The HIPP problem was shown by Lancia et al to be APX-hard~\cite{DBLP:journals/informs/LanciaPR04}.
\TODO{Need to mention what the complexity class APX is!}

@@

\section{SAT formulation}
\label{sec:hipp-sat}
\TODO{Clarify what each subscripted variable means to make reading easier.}

@@

Constraint (i) corresponds to equations $\sum_{1\leq k\leq r} s^a_{k,i} = 1$ and $\sum_{1\leq k\leq r} s^b_{k,i} = 1$ holding for all $i$ such that $1\leq i \leq n$.
This constraint is enforced in $\phi (\mathcal{R}, r)$ utilizing a propositional encoding of an \emph{1-bit adder circuit}.
The adder circuit uses variables $v^a_{k,i}$ where $1\leq k \leq r+1$ and $1\leq i\leq n$ and the following constraints (the ones for $s^b_{k,i}$ are similar).
\begin{eqnarray}
\forall i\text{ such that } 1\leq i\leq n: & v^a_{1,i} \leftrightarrow s^a_{1,i}\label{hipp-sat:clause1}\\
\forall k, i\text{ such that } 2\leq k \leq r,\ 1\leq i\leq n: & \neg v^a_{k,i} \vee \neg s^a_{k,i}\label{hipp-sat:clause2}\\
\forall k, i\text{ such that } 1\leq k \leq r,\ 1\leq i\leq n: & v^a_{k+1,i} \leftrightarrow (s^a_{k,i}\vee v^a_{k,i} )\label{hipp-sat:clause3}\\
\forall i\text{ such that } 1\leq i\leq n: & v^a_{r+1,i}=1\label{hipp-sat:clause4}
\end{eqnarray}
Constraint \eqref{hipp-sat:clause1} enforces $v^a_{1,i}$ and $s^a_{1,i}$ to take on the same value in every satisfying assignment whereas constraint \eqref{hipp-sat:clause4} enforces variable $v^a_{r+1,i}$ to be evaluated to true.
Constraint \eqref{hipp-sat:clause2} prevents $v^a_{k,i}$ and $s^a_{k,i}$ from being simultaneously assigned true, 
and constraint \eqref{hipp-sat:clause3} sets $v^a_{k+1,i}$ to 1 whenever either $s^a_{k,i}$ or $v^a_{k,i}$ is assigned 1.

@@

Next we formalize constraint (ii) stating that each genotype $g_i$ should be explained by its selected pair of haplotypes $(h_{k_1}, h_{k_2})$.
If $g_{i,j}$ (the $j$'th entry of $g_i$) is 0, then the $j$'th entries of the chosen haplotypes should also be 0.
This is encoded by $s^a_{k,i}\rightarrow\neg h_{k,j}$ and $s^b_{k,i}\rightarrow\neg h_{k,j}$ for all $1\leq k \leq r$.
Similarly, if $g_{i,j}$ is 1, then the $j$'th entries of the chosen haplotypes should also be 1, which is encoded by $s^a_{k,i}\rightarrow h_{k,j}$ and $s^b_{k,i}\rightarrow h_{k,j}$ for all $1\leq k \leq r$.
If $g_{i,j}$ is 2, then the $j$'th entry of the haplotypes should differ.
Variables $g^a_{i,j}$ and $g^b_{i,j}$ are used to decide which haplotype should be assigned 0 and which should be assigned 1.
The clauses $(g^a_{i,j}\vee g^b_{i,j})\wedge(\neg g^a_{i,j}\vee \neg g^b_{i,j})$ enforce $g^a_{i,j}$ and $g^b_{i,j}$ to take on different values.
The constraints $s^a_{k,i}\rightarrow (h_{k,j}\leftrightarrow g^a_{i,j})$ and $s^b_{k,i}\rightarrow (h_{k,j}\leftrightarrow g^b_{i,j})$ then enforce the selected haplotypes to take on values dictated by $g^a_{i,j}$ and $g^b_{i,j}$.
\begin{figure}
\centering
\small
\begin{mdframed}
\begin{align}
\text{for all  }& i\in\{1,\dots , n\}:\nonumber\\
& \quad\quad\quad\quad \sum_{1\leq k \leq r} s^a_{k,i} = 1 \\
& \quad\quad\quad\quad \sum_{1\leq k \leq r} s^b_{k,i} = 1 \\
\text{for all  }& i\in\{1,\dots , n\},\ k\in\{1,\dots r\}\text{ and }\nonumber\\&\quad\quad j\in\{1,\dots ,m\} \text{ such that }g_{i,j}=0:\nonumber\\
& \quad\quad\quad\quad s^a_{k,i}\rightarrow \neg h_{k,j} \\
& \quad\quad\quad\quad s^b_{k,i}\rightarrow \neg h_{k,j} \\
\text{for all  }& i\in\{1,\dots , n\},\ k\in\{1,\dots r\}\text{ and }\nonumber\\&\quad\quad j\in\{1,\dots ,m\} \text{ such that }g_{i,j}=1:\nonumber\\
& \quad\quad\quad\quad s^a_{k,i}\rightarrow h_{k,j} \\
& \quad\quad\quad\quad s^b_{k,i}\rightarrow h_{k,j} \\
\text{for all  }& i\in\{1,\dots , n\},\ k\in\{1,\dots r\}\text{ and }\nonumber\\&\quad\quad j\in\{1,\dots ,m\} \text{ such that }g_{i,j}=2:\nonumber\\
& \quad\quad\quad\quad g^a_{i,j}\vee g^b_{i,j} \\
& \quad\quad\quad\quad \neg g^a_{i,j}\vee \neg g^b_{i,j} \\
& \quad\quad\quad\quad s^a_{k,i}\rightarrow (h_{k,j}\leftrightarrow g^a_{i,j}) \\
& \quad\quad\quad\quad s^b_{k,i}\rightarrow (h_{k,j}\leftrightarrow g^b_{i,j}) \\
\end{align}
\end{mdframed}
\caption{The plain HIPP-SAT encoding.}
\label{fig:enc-hipp-sat}
\end{figure}

The encoding is presented in Figure~\ref{fig:enc-hipp-sat}.

Lynce and Marques-Silva use two techniques to facilitate and speed up the use of their HIPP-SAT encoding~\cite{DBLP:conf/aaai/LynceM06}.
There are symmetries present in the plain HIPP-SAT encoding, but fortunately these symmetries can be dispatched by utilizing symmetry-breaking predicates.
Since the usage of the encoding consists of calling $\phi (\mathcal{G}, r)$ with values of $r$ from some lower bound $lb$ up to $2n$ until the formula becomes satisfiable, the overall effort can be reduced by finding a large $lb$.
The computation of lower bounds of $r$ are thus useful in speeding up SAT-based haplotype inference.

@@

To compute the number of distinct haplotypes in the collection of $2n$ haplotypes variables $x_i$ with $1\leq i\leq 2n$ are included.
The constraints
$$ \text{for all }i\in\{1,\dots ,2n\}:\quad  x_{i} \geq 2 - i \sum_{j=1}^{i-1} d_{i,j} $$
ensure that $x_i$ is set to 1 whenever haplotype $h_i$ is distinct from haplotypes $h_1,\dots , h_{i-1}$.
The sum $\sum_{1\leq i\leq 2n} x_i$ then corresponds to the number of distinct haplotypes used.
The objective of the encoding is therefore to minimize $\sum_{1\leq i\leq 2n} x_i$.
The PolyIP encoding is presented in Figure~\ref{fig:enc-polyip}.

@@

We have now presented three encodings for solving the HIPP problem, one SAT encoding (HIPP-SAT) and two ILP encodings (RTIP and PolyIP).
These encoding differ both in terms of their structure as well as their size, performance and available optimizations.
The HIPP-SAT and PolyIP encodings are structurally similar since they both utilize a structured representation of the haplotypes thus giving the solvers access to this representation.
In contrast the RTIP encoding uses an atomic representation of the haplotypes and computes necessary information about the relationship of the haplotypes and genotypes in a pre-processing phase.

@@

The RTIP encoding is relatively straightforward since part of the work is done in the pre-processing phase.
Due to this fact the optimizations used for RTIP mostly concern the pre-processing phase.
Gusfield~\cite{DBLP:conf/cpm/Gusfield03} explains how the number of haplotypes to unfold can be reduced and how the unfolding can be accomplished efficiently.
The optimizations utilized in the HIPP-SAT and PolyIP formulations, on the other hand, modify the plain encoding itself by, e.g., including symmetry-breaking or transitivity constraints.

>>>>

\documentclass[12pt,a4paper]{article}
\usepackage[margin=1.5in]{geometry}

@@

\section{Introduction}
Haplotype inference is a crucial problem of modern genetics with wide-ranging applications in, e.g., correlating disease susceptibility to individuals' haplotypes.
Due to technical difficulties, the direct measurement of haplotypes is challenging, and current methods rely on indirectly inferring haplotypes from genotype data.

@@

In practice the input formula is usually restricted to be in \emph{conjunctive normal form} (CNF).
A formula $\phi$ is in CNF if it is a conjunction of \emph{clauses}
$$ \bigwedge_{i\in I} C_i $$
where each clause $C_i$ is a disjunction of \emph{literals}
$$ C_i = \bigvee_{j\in J_i} l_j $$
and each literal $l_j$ is either an atom $p$ or a negated atom $\neg p$.
CNF is a convenient standard format partly due to the fact that efficient polynomial-time algorithms exist for transforming arbitrary formulas into \emph{linear-size, equisatisfiable} CNF formulas~\cite{tseitin1983, DBLP:journals/jsc/PlaistedG86}.

@@

\section{Haplotype inference}
\label{sec:hipp}
The human genetic information is contained in the double-stranded DNA molecules consisting of a linear sequence of 4 bases (\textbf{A}denosine, \textbf{C}ytosine, \textbf{G}uanine, \textbf{T}hymine) in the phosphate-deoxyribose backbone.
The DNA molecules, which are packed into \emph{chromosomes}, consist of protein-coding intervals, which are called genes, and other non-coding parts.
Different variants of a gene are called \emph{alleles}, and there can exist 2 or more different alleles for a gene.
Humans are a \emph{diploid} species, which means that we have two copies of each chromosome (one from each parent).
We thus have two alleles of each gene, and we are said to be \emph{homozygous} with regard to a gene if we have two identical alleles and \emph{heterozygous} otherwise.

\emph{Haplotype} refers to the information contained in a single chromosome whereas the conflated information contained in the pair of  \emph{homologous} chromosomes inherited is referred to as the \emph{genotype}.
The genotype and haplotype data can take the form of full DNA sequences, \emph{microsatellite repetitions} or so-called \emph{single nucleotide polymorphisms} (SNPs).
Single nucleotide polymorphisms represent the variation at a single nucleotide on the population level, and usually this variation is mostly between two nucleotides.

In practice it is difficult and expensive to experimentally detect haplotype information.
Detecting genotype information has, however, become much faster, cheaper and more reliable in the recent decades.
Potential haplotypes can be computationally inferred from genotype data, and this is the problem referred to as the haplotype inference (HI) problem.
Haplotype data has the potential to yield new insight into genetic predisposition to, e.g., diseases, which makes the HI problem of great importance.

The HI problem formulations can be classified as \emph{biallelic} (two alleles) and \emph{multiallelic} (more than two alleles).
The following subsection considers the biallelic problem formulation motivated by the fact that most SNPs vary mostly between two nucleotides.

@@

This condition has been experimentally validated on both simulated and real data to yield relatively accurate haplotype data~\cite{DBLP:journals/bioinformatics/WangX03a}.
It has also been observed that the accuracy tends to increase with increasing sample size, i.e., the number $n$ of genotypes in the population~\cite{DBLP:journals/bioinformatics/WangX03a}.

The HIPP problem was shown by Lancia et al to be APX-hard~\cite{DBLP:journals/informs/LanciaPR04}.
This indicates that the HIPP problem is difficult to solve even approximately since APX is the complexity class of NP optimization problems that have a polynomial-time algorithm yielding an answer at most a constant factor away from the optimal solution~\cite{DBLP:books/lib/Ausiello99}.

@@

\section{SAT formulation}
\label{sec:hipp-sat}
This section presents a SAT encoding for solving the HIPP problem by Lynce and Marques-Silva~\cite{DBLP:conf/aaai/LynceM06}.
We refer to the encoding as HIPP-SAT.

@@

Constraint (i) corresponds to equations $\sum_{1\leq k\leq r} s^a_{k,i} = 1$ and $\sum_{1\leq k\leq r} s^b_{k,i} = 1$ holding for all $i$ such that $1\leq i \leq n$.
This constraint is enforced in $\phi (\mathcal{R}, r)$ utilizing a propositional encoding of a \emph{1-bit adder circuit}.
The adder circuit uses variables $v^a_{k,i}$ where $1\leq k \leq r+1$ and $1\leq i\leq n$ and the following constraints (the ones for $s^b_{k,i}$ are similar).
\begin{eqnarray}
\forall i\text{ such that } 1\leq i\leq n: & v^a_{1,i} \leftrightarrow s^a_{1,i}\label{hipp-sat:clause1}\\
\forall k, i\text{ such that } 2\leq k \leq r,\ 1\leq i\leq n: & \neg v^a_{k,i} \vee \neg s^a_{k,i}\label{hipp-sat:clause2}\\
\forall k, i\text{ such that } 1\leq k \leq r,\ 1\leq i\leq n: & v^a_{k+1,i} \leftrightarrow (s^a_{k,i}\vee v^a_{k,i} )\label{hipp-sat:clause3}\\
\forall i\text{ such that } 1\leq i\leq n: & v^a_{r+1,i}=1\label{hipp-sat:clause4}
\end{eqnarray}
Constraint \eqref{hipp-sat:clause1} enforces $v^a_{1,i}$ and $s^a_{1,i}$ to take on the same value in every satisfying assignment whereas constraint \eqref{hipp-sat:clause4} enforces variable $v^a_{r+1,i}$ to be evaluated to true.
Constraint \eqref{hipp-sat:clause2} prevents $v^a_{k,i}$ and $s^a_{k,i}$ from being simultaneously assigned true, 
and constraint \eqref{hipp-sat:clause3} sets $v^a_{k+1,i}$ to 1 whenever either $s^a_{k,i}$ or $v^a_{k,i}$ is assigned 1.

@@

Next we formalize constraint (ii) stating that each genotype $g_i$ should be explained by its selected pair of haplotypes $(h_{k_1}, h_{k_2})$.
If $g_{i,j}$ (the $j$'th entry of $g_i$) is 0, then the $j$'th entries of the chosen haplotypes should also be 0.
This is encoded by $s^a_{k,i}\rightarrow\neg h_{k,j}$ and $s^b_{k,i}\rightarrow\neg h_{k,j}$ for all $1\leq k \leq r$.
Similarly, if $g_{i,j}$ is 1, then the $j$'th entries of the chosen haplotypes should also be 1, which is encoded by $s^a_{k,i}\rightarrow h_{k,j}$ and $s^b_{k,i}\rightarrow h_{k,j}$ for all $1\leq k \leq r$.
If $g_{i,j}$ is 2, then the $j$'th entry of the haplotypes should differ.
Variables $g^a_{i,j}$ and $g^b_{i,j}$ are used to decide which haplotype should be assigned 0 and which should be assigned 1.
The clauses $(g^a_{i,j}\vee g^b_{i,j})\wedge(\neg g^a_{i,j}\vee \neg g^b_{i,j})$ enforce $g^a_{i,j}$ and $g^b_{i,j}$ to take on different values.
The constraints $s^a_{k,i}\rightarrow (h_{k,j}\leftrightarrow g^a_{i,j})$ and $s^b_{k,i}\rightarrow (h_{k,j}\leftrightarrow g^b_{i,j})$ then enforce the selected haplotypes to take on values dictated by $g^a_{i,j}$ and $g^b_{i,j}$.
\begin{figure}
\centering
\small
\begin{mdframed}
\begin{align}
\text{for all  }& i\in\{1,\dots , n\}:\nonumber\\
& \quad\quad\quad\quad \sum_{1\leq k \leq r} s^a_{k,i} = 1 \\
& \quad\quad\quad\quad \sum_{1\leq k \leq r} s^b_{k,i} = 1 \\
\text{for all  }& i\in\{1,\dots , n\},\ k\in\{1,\dots r\}\text{ and }\nonumber\\&\quad\quad j\in\{1,\dots ,m\} \text{ such that }g_{i,j}=0:\nonumber\\
& \quad\quad\quad\quad s^a_{k,i}\rightarrow \neg h_{k,j} \\
& \quad\quad\quad\quad s^b_{k,i}\rightarrow \neg h_{k,j} \\
\text{for all  }& i\in\{1,\dots , n\},\ k\in\{1,\dots r\}\text{ and }\nonumber\\&\quad\quad j\in\{1,\dots ,m\} \text{ such that }g_{i,j}=1:\nonumber\\
& \quad\quad\quad\quad s^a_{k,i}\rightarrow h_{k,j} \\
& \quad\quad\quad\quad s^b_{k,i}\rightarrow h_{k,j} \\
\text{for all  }& i\in\{1,\dots , n\},\ k\in\{1,\dots r\}\text{ and }\nonumber\\&\quad\quad j\in\{1,\dots ,m\} \text{ such that }g_{i,j}=2:\nonumber\\
& \quad\quad\quad\quad g^a_{i,j}\vee g^b_{i,j} \\
& \quad\quad\quad\quad \neg g^a_{i,j}\vee \neg g^b_{i,j} \\
& \quad\quad\quad\quad s^a_{k,i}\rightarrow (h_{k,j}\leftrightarrow g^a_{i,j}) \\
& \quad\quad\quad\quad s^b_{k,i}\rightarrow (h_{k,j}\leftrightarrow g^b_{i,j}) \\
\end{align}
\end{mdframed}
\caption{The plain HIPP-SAT encoding.}
\label{fig:enc-hipp-sat}
\end{figure}
The encoding is presented in Figure~\ref{fig:enc-hipp-sat}.

Lynce and Marques-Silva use two techniques to facilitate and speed up the use of their HIPP-SAT encoding~\cite{DBLP:conf/aaai/LynceM06}.
There are symmetries present in the plain HIPP-SAT encoding, but fortunately these symmetries can be dispatched by utilizing symmetry-breaking predicates.
Since the usage of the encoding consists of calling $\phi (\mathcal{G}, r)$ with values of $r$ from some lower bound $lb$ up to $2n$ or until the formula becomes satisfiable, the overall effort can be reduced by finding a large $lb$.
The computation of lower bounds of $r$ are thus useful in speeding up SAT-based haplotype inference.

@@

To compute the number of distinct haplotypes in the collection of $2n$ haplotypes variables $x_i$ with $1\leq i\leq 2n$ are included.
The constraints
$$ \text{for all }i\in\{1,\dots ,2n\}:\quad  x_{i} \geq 2 - i  + \sum_{j=1}^{i-1} d_{i,j} $$
ensure that $x_i$ is set to 1 whenever haplotype $h_i$ is distinct from haplotypes $h_1,\dots , h_{i-1}$.
The sum $\sum_{1\leq i\leq 2n} x_i$ then corresponds to the number of distinct haplotypes used.
The objective of the encoding is therefore to minimize $\sum_{1\leq i\leq 2n} x_i$.
The PolyIP encoding is presented in Figure~\ref{fig:enc-polyip}.

@@

Three encodings for solving the HIPP problem were presented in the preceeding sections; one SAT encoding (HIPP-SAT) and two ILP encodings (RTIP and PolyIP).
These encoding differ both in terms of their structure as well as their size, performance and available optimizations.
The HIPP-SAT and PolyIP encodings are structurally similar since they both utilize a structured representation of the haplotypes thus giving the solvers access to this representation.
In contrast the RTIP encoding uses an atomic representation of the haplotypes and computes necessary information about the relationship of the haplotypes and genotypes in a pre-processing phase.

@@

The RTIP encoding is relatively straightforward since a large part of the work is done in the pre-processing phase.
Due to this fact the optimizations used for RTIP mostly concern the pre-processing phase.
Gusfield~\cite{DBLP:conf/cpm/Gusfield03} explains how the number of haplotypes to unfold can be reduced and how the unfolding can be accomplished efficiently.
The optimizations utilized in the HIPP-SAT and PolyIP formulations, on the other hand, modify the plain encoding itself by, e.g., including symmetry-breaking or transitivity constraints.
